---
title: "Africa and its Urban Growth"
author: "Jan Sodoge"
date: "4/1/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

This markdown provides the data sources used and scripts applied to compute statistics and graphics used within the article on urbanization on the african continent. 
Data from Africapolis.com on urban growth.

```{r}
library(sf)
library(data.table)
library(raster)
library(dplyr)
library(gganimate)
library(spData)
library(ggridges)
library(tidyverse)
library(tmap)    # for static and interactive maps
library(leaflet) # for interactive maps
library(mapview) # for interactive maps
library(ggplot2) # tidyverse data visualization package
library(shiny)   # for web applications
library(rgdal)
library(readxl)
#data import
Africapolis_agglomeration_2015 <- read_excel("data/africa/Africapolis_agglomeration_2015.xlsx", 
                                             skip = 15) #<- datafile downloaded from Africapolis
remove_countries <- c("Spain", "Portugal", "Italy", "Greece",
                      "Saudi Arabia", "Israel", "Jordan",
                      "Yemen", "Oman", "Syria", "Iran", "Iraq",
                      "Turkey", "United Arab Emirates",
                      "Qatar", "Kuwait", "Cyprus",
                      "Azerbaijan", "Armenia",
                      "Turkmenistan")  
africa <- map_data("world")
africa <- dplyr::filter(africa, !(region  %in% remove_countries))


urban_data <- Africapolis_agglomeration_2015

```





# A first map
```{r}
    ggplot(data = africa, mapping = aes(x= long, y = lat,
       group = group))+
       geom_polygon(fill = "white", color = "black")+
       coord_map(projection = "mercator")+
       xlim(c(-20,60))+
       ylim(c(-40,40))+
       geom_point(data=Africapolis_agglomeration_2015, 
                  aes(x=Longitude, y=Latitude, group = NULL),size=0.1)+
      theme(plot.background = element_rect(fill = "#1D809F",
                         colour = "#1D809F",
                         size = 0.5, linetype = "solid"),
            panel.grid.major = element_blank(), 
            panel.grid.minor = element_blank(),
            panel.background = element_rect(fill = "#1D809F",
                         colour = "#1D809F",
                         size = 0.5, linetype = "solid"),
            axis.ticks = element_blank(),
            axis.text = element_blank()
          
            )+
      labs(x="",
           y="")

```


# Non-spatial Data Analysis
```{r}


million_cities_2015 <-  dplyr::filter(urban_data, Population_2015 > 1000000)
million_cities_2010 <-  dplyr::filter(urban_data, Population_2010 > 1000000)
million_cities_2000 <-  dplyr::filter(urban_data, Population_2000 > 1000000)
million_cities_1990 <-  dplyr::filter(urban_data, Population_1990 > 1000000)
million_cities_1980 <-  dplyr::filter(urban_data, Population_1980 > 1000000)
million_cities_1970 <-  dplyr::filter(urban_data, Population_1970 > 1000000)
million_cities_1960 <-  dplyr::filter(urban_data, Population_1960 > 1000000)
million_cities_1950 <-  dplyr::filter(urban_data, Population_1950 > 1000000)

million_cities_development <- data.table(c(nrow(million_cities_1950), nrow(million_cities_1960), nrow(million_cities_1970),
  nrow(million_cities_1980), nrow(million_cities_1990), nrow(million_cities_2000),
  nrow(million_cities_2010), nrow(million_cities_2015)), 
  c(1950, 1960, 1970, 1980, 1990, 2000, 2010, 2015))
 
```



```{r}

ggplot(data = million_cities_development, aes(x= V2, y = V1))+
  geom_line()+
  geom_point(color="#1D809F", size = 5)+
  theme(plot.background = element_rect(fill = "grey"),
            panel.grid.major = element_blank(), 
            panel.grid.minor = element_blank(),
            panel.background = element_rect(fill = "grey",
                         colour = "grey",
                         size = 0.5, linetype = "solid"),
            panel.grid.major.y = element_line(size=0.2, color = "white"))+
  labs(x = "",
       y = "Number of Megacities")+
  annotate("text", x = 1969, y = 23, label = "Alexandria and Cairo")+ #<- code to mark alexandria and cairo
    geom_curve(aes(x=1960, y=22, xend=1950, yend=6), color="black", size=0.2,
                     arrow = arrow(length = unit(0.03, "npc")))+
  annotate("rect", xmin = 1998, xmax = 2016, ymin = 22, ymax = 80,
  alpha = .2)+
  geom_curve(aes(x=1980, y=50, xend=2000, yend=60), color="black", size=0.2,
                     arrow = arrow(length = unit(0.03, "npc")))+
  annotate("text", x = 1970, y = 53, label = "44 new Megacities between 2000 and 2015")


```







```{r}
#analyzing the growth of africa´s current million inhabitants cities
megacities_growth <- million_cities_2015 %>% 
                     select(Agglomeration_Name, Population_1950, Population_1960, 
                              Population_1970, Population_1980, Population_1990,
                              Population_2000, Population_2010, Population_2015) %>% 
                    pivot_longer(-Agglomeration_Name, names_to = "citizens", values_to = "count") %>% 
                    mutate(citizens  = as.factor(citizens))



# compute the average growth rates of the megacities
growth_rates_megacities <- million_cities_2015 %>% 
  mutate(growth_period_1 = ((Population_1960 - Population_1950)/Population_1950)*100,
         growth_period_2 = ((Population_1970 - Population_1960)/Population_1960)*100,
         growth_period_3 = ((Population_1980 - Population_1970)/Population_1970)*100,
         growth_period_4 = ((Population_1990 - Population_1980)/Population_1980)*100,
         growth_period_5 = ((Population_2000 - Population_1990)/Population_1990)*100,
         growth_period_6 = ((Population_2010 - Population_2000)/Population_2000)*100) %>% 
         select(Agglomeration_Name, growth_period_1, growth_period_2, growth_period_3, growth_period_4,
                growth_period_5, growth_period_6) %>% 
        pivot_longer(-Agglomeration_Name, names_to = "citizens", values_to = "count") %>% 
        mutate(citizens  = as.factor(citizens))
        
         
         


```


```{r}
  ggplot(data = dplyr::filter(megacities_growth, Agglomeration_Name != "Cairo"), 
                              aes(x = citizens, y = count/1000, group = Agglomeration_Name))+
    geom_hline(yintercept = 1000, color = "white", size = 2)+ # <- the million citizens line

  geom_line(alpha=0.2)+
  geom_point(alpha=0.2, color="#1D809F")+
  scale_y_continuous(trans='log10')+
  ylim(c(10, 15000))+
  geom_curve(aes(y=5000, x=2, xend=0.7, yend=1100), color="black", size=0.2,
                     arrow = arrow(length = unit(0.03, "npc")))+
  annotate("text", x = 3.1, y = 5000, label = "Becoming a Megacity")+
  labs(x = "",
       y = "People [Tsd.]",
       title = "Africa´s Megacities Growth")+
   theme(plot.background = element_rect(fill = "grey"),
            panel.grid.major = element_blank(), 
            panel.grid.minor = element_blank(),
            panel.background = element_rect(fill = "grey",
                         colour = "grey",
                         size = 0.5, linetype = "solid"),
            panel.grid.major.y = element_line(size=0.2, color = "white"))+
  geom_curve(aes(y=11000, x=6, xend=8, yend=15000), color="black", size=0.2,
                     arrow = arrow(length = unit(0.03, "npc")))+
  annotate("text", x = 4, y = 11000, label = "Cairo somewhere up here: 23 Mil. in 2010")+
  scale_x_discrete(labels=c("Population_1950" = "1950",
                            "Population_1960" = "1960",
                            "Population_1970" = "1970",
                            "Population_1980" = "1980",
                            "Population_1990" = "1990",
                            "Population_2000" = "2000", 
                            "Population_2010" = "2000",
                            "Population_2015" = "2000"))
  
  
  
```

For the 74 megacities of 2015, the development of each is tracked to the left. While Cairo is excluded from the visualization (as it´s number of citizens exceeds a reasonable scale), we see most cities crossing the threshold of a million people between 198ß and 2000. At latest in 2000, all crossed the magic bar. 




```{r}
growth_rates_megacities_edited <- dplyr::filter(growth_rates_megacities, count < 100 & count > 0)  #<-filter for outliers
ggplot(data  = growth_rates_megacities_edited, aes(x = count, y = citizens))+
     geom_density_ridges(color="black", fill = "#1D809F", stat="binline", binwidth = 10)+
     labs(x = "Growth Rate",
          y = "",
          "Megacities Growth Rates")+
  scale_y_discrete(labels=c("growth_period_1" = "1950 to 1960",
                            "growth_period_2" = "1960 to 1970",
                            "growth_period_3" = "1970 to 1980",
                            "growth_period_4" = "1980 to 1990",
                            "growth_period_5" = "1990 to 2000",
                            "growth_period_6" = "2000 to2010"))+
  xlim(c(0,100))+
  theme(plot.background = element_rect(fill = "grey"),
            panel.grid.major = element_blank(), 
            panel.grid.minor = element_blank(),
            panel.background = element_rect(fill = "grey",
                         colour = "grey",
                         size = 0.5, linetype = "solid"),
            panel.grid.major.y = element_line(size=0.2, color = "white"))
  

```
 
 
 
 
 

```{r}

megacities_growth_edited <- left_join(megacities_growth, Africapolis_agglomeration_2015, by = "Agglomeration_Name")
    
megacities_growth_edited <- dplyr::filter(megacities_growth_edited, citizens == "Population_1950")

       ggplot(data = africa, mapping = aes(x= long, y = lat,
       group = group))+
       geom_polygon(fill = "white", color = "black")+
       coord_map(projection = "mercator")+
       xlim(c(-20,60))+
       ylim(c(-40,40))+
       geom_point(data = megacities_growth_edited, 
                  aes(x= Longitude, y=Latitude, group = NULL, size = count/100000),
                  fill = "green", pch=21,
                      color = "black", alpha = 0.5)+
      scale_size_identity(trans="sqrt",guide="legend")+
      theme(plot.background = element_rect(fill = "#1D809F",
                         colour = "#1D809F",
                         size = 0.5, linetype = "solid"),
            panel.grid.major = element_blank(), 
            panel.grid.minor = element_blank(),
            panel.background = element_rect(fill = "#1D809F",
                         colour = "#1D809F",
                         size = 0.5, linetype = "solid"),
            axis.ticks = element_blank(),
            axis.text = element_blank(),
            legend.background = element_rect(fill = "grey"),
            legend.key = element_rect(fill = "grey", color = "grey"),
            legend.position = "none" )+
        labs(x="",
             y="",
             title="1950")
   
```






```{r}

megacities_growth_edited <- left_join(megacities_growth, Africapolis_agglomeration_2015, by = "Agglomeration_Name")
    
megacities_growth_edited <- dplyr::filter(megacities_growth_edited, citizens == "Population_2015")

       ggplot(data = africa, mapping = aes(x= long, y = lat,
       group = group))+
       geom_polygon(fill = "white", color = "black")+
       coord_map(projection = "mercator")+
       xlim(c(-20,60))+
       ylim(c(-40,40))+
       geom_point(data = megacities_growth_edited, 
                  aes(x= Longitude, y=Latitude, group = NULL, size = count/1000),
                  fill = "green", pch=21,
                      color = "black", alpha = 0.5)+
      theme(plot.background = element_rect(fill = "#1D809F",
                         colour = "#1D809F",
                         size = 0.5, linetype = "solid"),
            panel.grid.major = element_blank(), 
            panel.grid.minor = element_blank(),
            panel.background = element_rect(fill = "#1D809F",
                         colour = "#1D809F",
                         size = 0.5, linetype = "solid"),
            axis.ticks = element_blank(),
            axis.text = element_blank(),
            legend.background = element_rect(fill = "grey"),
            legend.key = element_rect(fill = "grey", color = "grey"),
            legend.position = "right" )+
        labs(x="",
             y="",
             title="Urban populations by 2015")+
      guides(size=guide_legend(title="People [thou.]"))
         
```




```{r}
  ggplot(data = africa, mapping = aes(x= long, y = lat,
       group = group))+
       geom_polygon(fill = "white", color = "black")+
  geom_curve(aes(y=lagos_lat, x=lagos_long, xend=20, yend=20),
                    color="black", size=0.1,
                     arrow = arrow(length = unit(0.03, "npc")))+
       coord_map(projection = "mercator")+
       xlim(c(-20,60))+
       ylim(c(-40,40))
       
         
         
```





```{r}
# library
library(tidyverse)
library(viridis)
 data_1 <- million_cities_2015 %>% 
          select(Agglomeration_Name, Metropole, Population_1950, Population_1960, Population_1970,
                 Population_1980, Population_1990, Population_2000, Population_2010) %>% 
          slice(1:10)
# Transform data in a tidy format (long format)
data <- data %>% gather(key = "observation", value="value", -c(1,2)) 
 
data <- data_1 %>% gather(key = "observation", value="value", -c(1,2)) %>% 
                   mutate(group = Metropole)

data$Metropole <- NULL
# Set a number of 'empty bar' to add at the end of each group
empty_bar <- 2
nObsType <- nlevels(as.factor(data$observation))
to_add <- data.frame( matrix(NA, empty_bar*nlevels(data$group)*nObsType, ncol(data)) )
colnames(to_add) <- colnames(data)
to_add$group <- rep(levels(data$group), each=empty_bar*nObsType )
data <- rbind(data, to_add)
data <- data %>% arrange(group, Agglomeration_Name)
data$id <- rep( seq(1, nrow(data)/nObsType) , each=nObsType)
 
# Get the name and the y position of each label
label_data <- data %>% group_by(id, Agglomeration_Name) %>% summarize(tot=sum(value))
number_of_bar <- nrow(label_data)
angle <- 90 - 360 * (label_data$id-0.5) /number_of_bar     # I substract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
label_data$hjust <- ifelse( angle < -90, 1, 0)
label_data$angle <- ifelse(angle < -90, angle+180, angle)
 
# prepare a data frame for base lines
base_data <- data %>% 
  group_by(group) %>% 
  summarize(start=min(id), end=max(id) - empty_bar) %>% 
  rowwise() %>% 
  mutate(title=mean(c(start, end)))
 
# prepare a data frame for grid (scales)
grid_data <- base_data
grid_data$end <- grid_data$end[ c( nrow(grid_data), 1:nrow(grid_data)-1)] + 1
grid_data$start <- grid_data$start - 1
grid_data <- grid_data[-1,]
 
# Make the plot
  ggplot(data) +      
  
  # Add the stacked bar
  geom_bar(aes(x=as.factor(reorder(id, value)), y=value, fill=observation), stat="identity", alpha=0.5) +
  scale_fill_viridis(discrete=TRUE) +
  

  
  ylim(-150,max(label_data$tot, na.rm=T)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(-1,4), "cm") 
  ) +
  coord_polar() 
  

```




```{r}
# library
library(tidyverse)
library(viridis)
 data_1 <- million_cities_2015 %>% 
          select(Agglomeration_Name, Metropole, Population_1950,
                 Population_1960, Population_1970, Population_1980,
                 Population_1990, Population_2000, Population_2010) %>% 
          slice(1:10)
# Transform data in a tidy format (long format)
data <- data %>% gather(key = "observation", value="value", -c(1,2)) 
 
data <- data_1 %>% gather(key = "observation", value="value", -c(1,2)) %>% 
                   mutate(group = Metropole)

data$Metropole <- NULL
# Set a number of 'empty bar' to add at the end of each group
empty_bar <- 2
nObsType <- nlevels(as.factor(data$observation))
to_add <- data.frame( matrix(NA, empty_bar*nlevels(data$group)*nObsType, ncol(data)) )
colnames(to_add) <- colnames(data)
to_add$group <- rep(levels(data$group), each=empty_bar*nObsType )
data <- rbind(data, to_add)
data <- data %>% arrange(group, Agglomeration_Name)
data$id <- rep( seq(1, nrow(data)/nObsType) , each=nObsType)
 
# Get the name and the y position of each label
label_data <- data %>% group_by(id, Agglomeration_Name) %>% summarize(tot=sum(value))
number_of_bar <- nrow(label_data)
angle <- 90 - 360 * (label_data$id-0.5) /number_of_bar     # I substract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
label_data$hjust <- ifelse( angle < -90, 1, 0)
label_data$angle <- ifelse(angle < -90, angle+180, angle)
 
# prepare a data frame for base lines
base_data <- data %>% 
  group_by(group) %>% 
  summarize(start=min(id), end=max(id) - empty_bar) %>% 
  rowwise() %>% 
  mutate(title=mean(c(start, end)))
 
# prepare a data frame for grid (scales)
grid_data <- base_data
grid_data$end <- grid_data$end[ c( nrow(grid_data), 1:nrow(grid_data)-1)] + 1
grid_data$start <- grid_data$start - 1
grid_data <- grid_data[-1,]
 
# Make the plot
  ggplot(data) +      
  
  # Add the stacked bar
  geom_bar(aes(x=as.factor(reorder(id, value)), y=value, fill=observation), stat="identity", alpha=0.5) +
  scale_fill_viridis(discrete=TRUE) +
  ylim(c(-150, 92995802))+
  

  
  
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(-1,4), "cm") 
  ) +
  coord_polar() 
  

```

